---
title: '**R** basics for beginners'
author: "Dae-Jin Lee"
date: "<http://idaejin.github.io/bcam-courses>"
output:
  slidy_presentation:
    fig_caption: yes
    highlight: haddock
    keep_md: yes
  ioslides_presentation:
    highlight: kate
    incremental: no
subtitle: Basque Center for Applied Mathematics
---



---------------------------------------

# Why R ? 

* **R** is a command-driven statistical package.

* The most important reasons to use **R** are:

    + **R** is free and multiplatform (Windows/Linux/MACos)

    + **R** allows you to do all the statistical tests/models/analysis you need :)

    + Excellent graphics and programming capabilities

    + Growing community of users and developers 

    + Lots of online resources

    + An user-friendly interface is RStudio <http://www.rstudio.com/>

-------------------------------

# Statistical features

* Graphical Techniques (Exploratory Data Analysis)
* Linear and non-linear modeling (linear regression, non-parametric regression, smoothing, etc ...)
* Classical statistical tests
* Time-series analysis
* Econometrics
* Classification and clustering (data mining, machine learning)
* Optimization and Mathematical Programming
* Bayesian inference etc ....

**Visit** http://cran.r-project.org/web/views
 or http://stackoverflow.com/questions/tagged/r
 
 
------------------------------
 
# Start with **R**

* Get current working directory 
```{r,eval=FALSE}
getwd() 
```

* Set working directory
```{r, eval=FALSE}
setwd("/Users/dlee") 
```

## Install and load an **R** library
 
```{r,eval=FALSE}
install.packages("DAAG") # (Data Analysis And Graphics)
```

* Once installed the package, load it
```{r,warning=FALSE,message=FALSE}
library(DAAG) # or require(DAAG)
```
 

-------------------------------------


# Data Import

* Several formats are available (`.txt`, `.csv`, `.xls`, `.xlsx`, `SAS`, `Stata`, etc...)
* Some **R** libraries to import data are 
```{r,message=FALSE,warning=FALSE}
library(gdata)
library(foreign)
``` 

\bigskip

* Read data from a `.txt` or `.csv` files (*e.g.:* created in `Rintro1.R`)
```{r,eval=FALSE}
mydata1 = read.table("cardata.txt") 
mydata2 = read.csv("cardata.csv")  
```

* Other formats `.xls` and `.xlsx`
\medskip

```{r,eval=FALSE,message=FALSE,warning=FALSE}
# read in the worksheet named mysheet
mydata <- read.xlsx("myexcel.xlsx", sheetName = "mysheet")
```

---------------------------------


* Minitab, SPSS, SAS or Stata \medskip

```{r, eval=FALSE, message=FALSE}
library(foreign)                   
mydata = read.mtp("mydata.mtp")  # Minitab
mydata = read.spss("myfile", to.data.frame=TRUE) # SPSS
mydata = read.dta("mydata.dta") # Stata
```

* Or
```{r,eval=FALSE}
library(Hmisc)
mydata = spss.get("mydata.por", use.value.labels=TRUE)  # SPSS
```

-------------------------

# Exporting data

* There are numerous methods for exporting **R** objects into other formats . For SPSS, SAS and Stata. you will need to load the `foreign` packages. For Excel, you will need the `xlsx` package.  
 
 - Tab delimited text file

```{r,eval=FALSE}
write.table(mydata, "mydata.txt", sep="\t") 
```

*  Excel spreadsheet

```{r,eval=FALSE}
library(xlsx)
write.xlsx(mydata, "mydata.xlsx")
```


----------------------------

# Data vectors

* Download `R code` [here](http://idaejin.github.io/bcam-courses/rbasics/rbasics.R)

* Create a vector of weights and heights

```{r}
weight<-c(60,72,57,90,95,72) # function c is used to concatenate data/numbers or vectors 
class(weight)
height<-c(1.75,1.80,1.65,1.90,1.74,1.91)
```

* calculate Body Mass Index
```{r} 
bmi<- weight/height^2
bmi
```

----------------------------


# Basic statistics 

* mean, median, st dev, variance

```{r,eval=FALSE}
mean(weight) 
median(weight)
sd(weight)
var(weight)
```

* summarize data

```{r}
summary(weight)
```

* or
```{r,eval=FALSE}
min(weight)
max(weight)
range(weight)
quantile(weight)
sum(weight)
length(weight)
```


----------------------------------

# Character vectors and factor variables

```{r,fig.width=5,fig.height=5}
subject <- c("John","Peter","Chris","Tony","Mary","Jane")
sex <- c("MALE","MALE","MALE","MALE","FEMALE","FEMALE")
class(subject)
table(sex)
```


-----------------------------------


# Data frames
```{r}
Dat <- data.frame(subject,sex,weight,height)
# add bmi to Dat
Dat$bmi <- bmi  # or Dat$bmi <- weight/height^2
class(Dat)
str(Dat) # display object structure
```


```{r}
# Change rownames
rownames(Dat)<-c("A","B","C","D","E","F")

# Access to data frame elements (similar to a matrix)
Dat[,1]     # 1st column
Dat[,1:3]   # 1st to 3rd columns
Dat[1:2,]   # 1st to 2nd row
```



# Working with data frames
## Example: Analyze data by groups

*  Obtain the mean weight, heigth and bmi means by FEMALES and MALES:

1. Select each group and compute the mean

```{r, results='hide'}
Dat[sex=="MALE",]
Dat[sex=="FEMALE",]

mean(Dat[sex=="MALE",3])  # weight average of MALEs
mean(Dat[sex=="MALE","weight"])
```

2. Use `apply` by columns 
```{r, results='hide'}
apply(Dat[sex=="FEMALE",3:5],2,mean)
apply(Dat[sex=="MALE",3:5],2,mean)

# we can use apply with our own function
apply(Dat[sex=="FEMALE",3:5],2,function(x){x+2})
```

3. `by` and `colMeans`
```{r, results='hide'}
by(Dat[,3:5],sex, colMeans) # 'by' splits your data by factors and do calculations on each subset.
```

4. `aggregate`

```{r, results='hide'}
# another option
aggregate(Dat[,3:5], by=list(sex),mean) 
```

------------------------------------

# Logical vectors

* Choose individuals with BMI>22
```{r,results='hide'}
bmi
bmi>22
as.numeric(bmi>22) # convert a logical condition to a numeric value 0/1
which(bmi>22)  # gives the position of bmi for which bmi>22
```

* Which are between 20 and 25?

```{r,results='hide'}
bmi > 20 & bmi < 25
which(bmi > 20 & bmi < 25)
```

--------------------------------------

# Working with vectors 

* Concatenate

```{r,results='hide'}
x <- c(2, 3, 5, 2, 7, 1)
y <- c(10, 15, 12)
z <- c(x,y)  # concatenates x and y
```

* list two vectors

```{r}
zz <- list(x,y) # create a list
unlist(zz) # unlist the list converting it to a concatenated vector
``` 

* subset of vectors

```{r}
x[c(1,3,4)]

x[-c(2,6)] # negative subscripts omit the chosen elements 
```

* Sequences
```{r}
seq(1,9) # or 1:9
seq(1,9,by=1)
seq(1,9,by=0.5)
seq(1,9,length=20)
```

* Replicates

```{r,results='hide'}
oops <- c(7,9,13)
rep(oops,3) # repeats the entire vector "oops" three times
rep(oops,1:3) # this function has the number 3 replaced 
              #  by a vector with the three values (1,2,3) 
              #  indicating that 7 should be repeated once, 9 twice and 13 three times.

rep(c(2,3,5), 4)
rep(1:2,c(10,15))

rep(c("MALE","FEMALE"),c(4,2)) # it also works with character vectors 
c(rep("MALE",3), rep("FEMALE",2))
```

---------------------------------------------


# Matrices and arrays

```{r}
x<- 1:12
x
dim(x)<-c(3,4)  # 3 rows and 4 columns

X <- matrix(1:12,nrow=3,byrow=TRUE)
X <- matrix(1:12,nrow=3,byrow=FALSE)

# rownames, colnames

rownames(X) <- c("A","B","C")
colnames(X) <- LETTERS[4:7]
colnames(X) <- month.abb[4:7]
```

* Column/Row bind operations `cbind()`, `rbind()`

```{r}
Y <- matrix(0.1*(1:12),3,4)

cbind(X,Y)  # bind column-wise
rbind(X,Y)  # bind row-wise
```

-----------------------------------------------

# Factors

```{r}
gender<-c(rep("female",691),rep("male",692))
class(gender)

# change vector to factor (i.e. a category)
gender<- factor(gender)
levels(gender)

summary(gender)
table(gender)

status<- c(0,3,2,1,4,5)    # This command creates a numerical vector ???pain???, encoding the pain level of five patients.
fstatus <- factor(status, levels=0:5)
levels(fstatus) <- c("student","engineer","unemployed","lawyer","economist","dentist")

Dat$status <- fstatus
Dat
````


-----------------------------------------------

#  Indexing vector with logicals

```{r}
a <- c(1,2,3,4,5)
b <- c(TRUE,FALSE,FALSE,TRUE,FALSE)

max(a[b])

sum(a[b])
````

# Missing values (NA)

```{r}
a <- c(1,2,3,4,NA)
sum(a)
sum(a,na.rm=TRUE)

a <- c(1,2,3,4,NA)
is.na(a)
```



------------------------------------------------

# Working with data frames

 * A data frame is used for storing data tables. It is a list of vectors of equal length. 
```{r,results="hide"}
mtcars
?mtcars       # or help(mtcars)
```

* look at the first rows

```{r}
head(mtcars)
```

* Structure of the data frame

```{r}
str(mtcars) # display the structure of the data frame
```

* Select a car model:
```{r}
mtcars["Mazda RX4",] # using rows and columns names
mtcars[c("Datsun 710", "Camaro Z28"),] 
```

* Or specific variables

```{r}
mtcars[,c("mpg","am")]
```

------------------------------------

# Plotting 

*  Scatterplot

```{r}
attach(mtcars)
plot(wt, mpg, main="Scatterplot Example",
   xlab="Car Weight ", ylab="Miles Per Gallon ", pch=19) 
```

* Basic Scatterplot Matrix
```{r}
pairs(~mpg+disp+drat+wt,data=mtcars,
   main="Simple Scatterplot Matrix")
```

* Barplot

```{r,fig.pos='center'}
tab <- table(mtcars[,c("cyl")])
barplot(tab)
```

* Piechart

```{r}
pie(tab)
```



-----------------------------------------

# Exercises

1. The data.frame `VADeaths` contains the death rates per 1000 in Virginia (US) in 1940

    + The death rates are measured per 1000 population per year. They are cross-classified by age group (rows) and population group (columns). The age groups are: 50–54, 55–59, 60–64, 65–69, 70–74 and the population groups are Rural/Male, Rural/Female, Urban/Male and Urban/Female.

```{r}
VADeaths
```

* Compute the mean for each age group. 
      + **Result:**

```{r,echo=FALSE}
apply(VADeaths,1,mean)
```

* Compute the mean for each population group. 
      + **Result:** 
      
```{r,echo=FALSE}
apply(VADeaths,2,mean)
```

2. The  `data.frame` `rainforest` contains several variables from different `species`

```{r, results='hide'}
library(DAAG)
rainforest
```

  * Create a table of counts for each `species` and make a graphic with the results. 
  
      + **Result:**
      
```{r,echo=FALSE}
table(rainforest$species)
barplot(table(rainforest$species))
```

3. The `Acmena` `data.frame` is created from  `rainforest` using the function `subset`. 

  * Plot the relationship between the wood biomass (`wood`) and the diameter of the breast height (`dbh`). Use also a logarithm scale.

```{r}
Acmena <- subset(rainforest, species == "Acmena smithii")
```

 

```{r,echo=FALSE}
par(mfrow=c(1,2))
plot(wood~dbh,data=Acmena,pch=19, main="plot of dbh vs wood")
plot(log(wood)~log(dbh),data=Acmena,pch=19,main="log transformation")
```


  * Compute a histogram of variable `dbh` using function `hist`

```{r,echo=FALSE}
hist(Acmena$dbh)
```

4. Create a vector of the positive odd integers less than 100 and remove the values greater than 60 and less than 80.

    * **Result:** 
    
```{r,echo=FALSE}
  x <- seq(1,100,by=2)
  x[x>60 & x<80]
```


* [Solutions here](http://idaejin.github.io/bcam-courses/rbasics/rbasics_sol.R)


```{r,eval=FALSE,echo=FALSE}
library(knitr)
purl("RBasics.Rmd",output="RBasics.R")
```






